Exponential Random Graph Models with Big Networks: Maximum Pseudolikelihood Estimation and the Parametric Bootstrap

### 2017-07-28 TeX/conference_paper.tex
finished revisions
finalized paper

\begin{abstract}
\noindent With the growth of interest in network data across fields, the Exponential Random Graph Model (ERGM) has emerged as the leading approach to the statistical analysis of network data. ERGM parameter estimation requires the approximation of an intractable normalizing constant. Simulation methods represent the state-of-the-art approach to approximating the normalizing constant, leading to estimation by Monte Carlo maximum likelihood (MCMLE). MCMLE is accurate when a large sample of networks is used to approximate the normalizing constant. However, MCMLE is computationally expensive, and may be prohibitively so if the size of the network is on the order of 1,000 nodes (i.e., one million potential ties) or greater. When the network is large, one option is maximum pseudolikelihood estimation (MPLE). The standard MPLE is simple and fast, but generally underestimates standard errors. We show that a resampling method---the parametric bootstrap---results in accurate coverage probabilities for confidence intervals. We find that bootstrapped MPLE can be run in 1/5th the time of MCMLE. We study the relative performance of MCMLE and MPLE with simulation studies, and illustrate the two different approaches by applying them to a network of bills introduced in the United State Senate.

@@

\section{Introduction}
\noindent The field of network science faces a double-edge sword when it comes to computationally intensive research. First, the availability of digital source data has led the growth in network science to be synonymous with the growth in research on big data.  Faraj et al. (2008, pp. 19--20) \cite{faraj2008electronic} notes that, \begin{quote} ``In recent years the proliferation of advanced Information Technology has not only facilitated the collection of large-scale network data but also increased the availability of large-scale network data analysis tools and techniques. This has led to an explosion of interest in large network research in fields spanning biology, physics, mathematics, and social sciences.''  \end{quote} Second, analytical methods are growing more sophisticated, increasingly involving iterative and/or simulation-based optimization, rather than simple descriptive calculations \cite{snijders2011statistical}.  Closing the gap in terms of the size of the networks to which it is feasible to apply the most sophisticated methods of network modeling requires research into scalable methods of inference. We propose a method of statistical inference for one of the most popular models for networks---the exponential random graph model (ERGM), in which both parameter estimates and confidence intervals are derived, that can require less than half the compute time of currently used methods.

\section{The Exponential Random Graph Model}
\noindent The ERGM is a probabilistic model for networks \cite{chatterjee2013estimating,Wasserman.1996,robins.pattison.2007}. They can be used for link prediction \cite{lu2010supervised}, simulating network adjacency matrices \cite{hackney2006agent}, and testing theories regarding the processes underlying tie formation \cite{goodreau2009birds}. The ERGM was first introduced by Holland and Leinhardt (1981) \cite{holland1981exponential}. However, due to the intractable normalizing constant in the likelihood function of the ERGM, it did not see widespread and complete use until the 2000s, following the development of algorithms and software for efficient simulation-based methods for working with ERGM \cite{snijders2002markov}. Training ERGM using simulation-based methods is computationally expensive, and can still be prohibitively burdensome with data on big networks. Approximate methods of estimation, which are much more feasible with large networks, have existed for some time, but these methods perform poorly when it comes to characterizing the uncertainty in parameter estimates, which is necessary when assessing risk in predictions or simulation, or in hypothesis testing.

The ERGM takes the adjacency matrix of an observed network $G^{obs}$, which is a matrix-valued random variable. This means that a network of $N$ nodes can be defined as a adjacency matrix $G=(g_{ij})\in \mathbb{[0,1]}^{N \times N}$, where $g_{ij} \in \{0,1\}$ for all $i,j \in \{1,\dots , N\}$. $g_{ij}=1$ means that there is an edge between actors $i$ and $j$, while $g_{ij}=0$ indicates that these actors are not directly connected. Since the model does not consider loops, one has $g_{ii}=0$ for all $i \in \{1,\dots , N\}$. Furthermore, define
$$ \mathcal{G}(N) := \left\{ G \in \mathbb{R}^{(N \times N)}: g_{ij} \in \{0,1\},~g_{ii}=0\right\}$$
as the set of all possible networks on $N$ nodes without loops. Note that the cardinality of set $\mathcal{G}(N)$ is increasing exponentially for every newly included actor, which results in $2^{N(N-1)/2}$ total elements. For even a small number of nodes, the cardinality of $\mathcal{G}(N)$ turns out to be an astronomically large number. For this reason calculating the likelihood function of the ERGM, which requires evaluating a normalizing constant on $\mathcal{G}(N)$  is either extremely time-consuming or with today's technology not achievable. As a consequence, many approximation methods have been provided by the literature, with the most popular method making use of Markov Chain Monte Carlo (MCMC) methods \cite{hunter2012computational}, as we will introduce in the next section.\\

The probability function for the ERGM is defined as
\begin{equation}
\mathbb{P}_{\theta}(G)=\dfrac{\exp(\theta^T \cdot \Gamma(G))}{\sum_{G^* \in \mathcal{G}(N)} \exp(\theta^T \cdot \Gamma(G^*))} 
\label{ERGM}
\end{equation}
%
where $\theta \in \mathbb{R}^q$ is a $q-$dimensional vector of parameters, $\Gamma:\mathcal{G}(N) \to \mathbb{R}^q~,~G \mapsto (\Gamma_1(G),\dots,\Gamma_q(G))^T$ is a $q$-dimensional function of different network statistics and $c(\theta):= \sum_{G^* \in \mathcal{G}(N)} \exp(\theta^T \cdot \Gamma(G^*))$ is a normalization constant which ensures that (\ref{ERGM}) defines a probability function on $\mathcal{G}$.
%
As already mentioned, a specific network $G$ can be considered as a manifestation of a matrix-like random variable, whose probability of occurrence can be modeled with equation (\ref{ERGM}) . The generative processes captured by a model (e.g., density, reciprocity, popularity, clustering) are informed by the decision regarding which network statistics (i.e., $\Gamma(\cdot)$) are incorporated (see Snijders et al. (2006) \cite{SnijdersTomA.B..2006} for a detailed discussion). The flexibility of the ERGM in capturing virtually any network generative process has led to it being applied broadly across several fields, including (but not limited to) sociology \cite{smith2016ethnic,srivastava2011culture}, economics \cite{lomi2012networks}, political science \cite{Cranmer.2011,lazer2010coevolution}, ecology \cite{dey2014individual,bouskila2015similarity}, and neuroscience \cite{simpson2011exponential,simpson2012exponential}.

\section{Estimation}
\noindent As mentioned above calculating $c(\theta)$ is not achievable for most cases with today's technology.  Straightforward application of either maximum likelihood estimation and Bayesian inference are, therefore, not possible. The first method proposed in the literature for estimating ERGM parameters was maximum pseudolikelihood estimation \cite{StraussIkeda1990}. Under maximum pseudolikelihood estimation (MPLE), the joint distribution is replaced by the product over conditional distributions \cite{besag1986statistical}.  
%
\begin{equation*}
\mathbb{P}_{\theta}(G) \approx \prod_{ij} \mathbb{P}_{\theta}\left(g_{ij}|G_{-ij}\right), 
%\label{ILogit}
\end{equation*}
where $G_{-ij}$ is the adjacency matrix, excluding element $ij$. The conditional probability of a tie in ERGM reduces, conveniently, to a logistic regression form given by 
\begin{equation*}
\mathbb{P}_{\theta}\left(g_{ij}=1| G_{-ij} \right) = \text{\\} \text{logit}^{-1}\left( \theta^T \cdot \delta(\Gamma(G))  \right), 
%\label{ILogit}
\end{equation*}
where $\delta(\Gamma(G))$ is the ``change statistic'' given by the difference in the network statistics when the $ij$ element is toggled from 0 to 1 (i.e., $\Gamma(G|g_{ij}=1))-\Gamma(G|g_{ij}=0)$), and $\text{logit}^{-1}(x)=1/(1+\exp(-x))$ \cite{goodreau2009birds}. For the ERGM, the pseudolikelihood function can be maximized using logistic regression software, in which the dependent variable is given by the elements of the adjacency matrix, and the covariates are given by the values of the change statistics corresponding to each element of the adjacency matrix.

@@

The contemporary conventional approach to estimating $\theta$, introduced by Snijders (2002) \cite{snijders2002markov}, is based on a Markov Chain Monte Carlo (MCMC) approximation of the MLE.\\[0.3cm]
This Monte Carlo maximum likelihood method (MCMLE) is based on a more direct attempt to approximate the  log-likelihood function derived from (\ref{ERGM}). The log-likelihood function is not evaluated directly, rather, the log ratio of the likelihood under a proposed value of the parameters $\theta$, and an initial value of the parameters $\theta_0$, is approximated using $L$ networks simulated from the ERGM with parameter values $\theta_0$. The approximation, detailed in Snijders (2002) \cite{snijders2002markov} is given by 
%
\begin{eqnarray*}
&&\text{loglik}(\theta)-\text{loglik}(\theta_0)\\&=&- \log(c(\theta))+\log(c(\theta_0))\\
                                             &=&- \log \left( \frac{c(\theta)}{c(\theta_0)} \right)\\
                                             &=&- \log \left( \mathbb{E}_{\theta_0}\left[ \exp\left((\theta - \theta_0)^T \cdot \Gamma(Y)\right) \right] \right)\\
                                             &\approx &- \log \left( \frac{1}{L} \cdot \sum_{i=1}^{L}  \exp \left((\theta - \theta_0)^T \cdot \Gamma(G_i) \right) \right).
\end{eqnarray*}
By differentiating this equation on both sides with respect to $\theta$ one gets an approximate score function:
%
\begin{equation}
s(\theta) \approx -\frac{\partial}{\partial \theta} \log \left( \frac{1}{L} \cdot \sum_{i=1}^{L}  \exp \left((\theta - \theta_0)^T \cdot \Gamma(G_i) \right) \right)
\label{score}
\end{equation}
%
This approximate score function now can be used as usual, i.e., it can be iteratively approximately optimized with the \textit{Newton-Raphson algorithm} or by \textit{Fisher scoring}.

MCMC methods are used to simulate the $L$ networks. As we demonstrate below, the MCMLE grows more accurate as $L$ increases. Indeed, MCMLE approaches the MLE as the number of networks simulated goes to infinity. Snijders (2002) \cite{Snijders.2002} provides a Metropolis Hastings algorithm to simulate networks:
Choose a matrix $G^{(0)} \in \mathcal{G}(N)$ to start with (e.g., the observed network). For $k \in \{0,...,L-1\}$ recursively proceed as follows:\\
\begin{enumerate}
\item Randomly choose an edge $(i,j)$ where $i \neq j$ from $G^{(k)}$
\item Compute the value
$$\pi := \dfrac{\mathbb{P}_{\theta}(Y_{ij} \neq g_{ij}^{(k)} | Y_{ij}^c=G_{ij}^c)}{\mathbb{P}_{\theta}(Y_{ij}= g_{ij}^{(k)} | Y_{ij}^c=G_{ij}^c)}$$
\item Fix $\delta:= \min\{1, \pi\}$ and draw a random number $Z$ from Bin$(1, \delta)$. If
\begin{itemize}
\item $Z=0$, let $G^{(k+1)} := G^{(k)}$ 
\item $Z=1$, define $G^{(k+1)}$ via
$$g_{pq}^{(k+1)}=\begin{cases}
1-g_{pq}^{(k)}& \text{if}~ (p,q)=(i,j) \\
g_{pq}^{(k)} &\text{if}~ (p,q) \neq (i,j) 
\end{cases}$$
\end{itemize}
\item Start at step 1 with $G^{(k+1)}$.
\end{enumerate}
\vspace{0.3cm}
\noindent The depicted algorithm provides a sequence of random networks $G_(0),\dots,G_(L)$ via a Metropolis-Hastings sampler \cite{chib1995understanding}. Since the original matrix was chosen randomly and the first simulated networks are very dependent on the chosen matrix (only one edge is changed per iteration), usually
the first $B$ networks, where $N << B << L$, are discarded as the so called Burn-In.

\section{Efficiency of MPLE and MCMLE}
\noindent  As mentioned in the previous section the MPLE approaches the MLE as the size of the networks increase and as a consequence, is a consistent estimator (see Lindsay \cite{Lindsay1988}, Strauss and Ikeda \cite{StraussIkeda1990}, Hyvarinen \cite{Hyvarinen2006}, Desmarais and Cranmer \cite{Desmarais.2012}). This implies that for an increasing number of nodes, the MPLE converges in probability to the MLE, meaning that for large enough networks the MPLE performs as well as or better than MCMLE, and requires less compute time. At this point we want to mention that we are familiar with the work of Shalizi and Rinaldo \cite{shalizi2013}, arguing that consistency is not given in the ERGM framework. They prove that one cannot run an ERGM on a sub-network in order to make inferences about the full network.  The way we use the term \textit{consistency} in this paper is different and aligns with the way consistency is defined by Lindsay \cite{Lindsay1988}, i.e. instead of considering sub-networks that converge to the full size network, we argue that both, the MLE as well as the MPLE, approach the true coefficient values as the size of networks generally increases.

To illustrate the relative efficiency of MPLE and MCMLE Next, we run a simulation study. Desmarais and Cranmer \cite{Desmarais.2012} show the MPLE outperforms the MCMLE if the number of simulated networks used to approximate the likelihood in MCMLE is not large enough. It is even more remarkable that the number of simulated networks needed for the MCMLE, in order to surpass the MPLE increases as the size of the network (i.e., the number of nodes in the network) increases. This means that, for very large networks, it becomes difficult to determine the number of simulated networks required for the MCMLE to outperform the MPLE. In other words, the larger the network of the more computationally intensive it becomes to use MCMLE in a way that out-performs MPLE.\\[0.3cm]
\noindent To demonstrate this disadvantage of the MCMLE we conduct a simulation study using Goodreau's \cite{HunterDavidR..2008} Faux Mesa High School data, which represents a simulation of an in-school friendship network among 203 students as well as the Faux Magnolia High School data, representing an in-school friendship network among 1451 students. The data for both networks originates from Resnick et al. \cite{Resnicketal1997}.\\[0.3cm]
For both networks, we first calculate the MCMLE and treat the estimated coefficients as the network's true values $\theta$. Then, we take the same parametrization, using the number of edges, the nodal attribute for gender, and the geometrically weighted edgewise shared partners (gwesp) distribution (see Hunter \cite{Hunter.2006}) where we fix the decay parameter $\lambda$ at $0.25$.

@@

We simulate $m=500$ new networks using the 'true' coefficients and estimate the MPLE as well as the MCMLE of these simulated networks. For every single simulated network the MCMLE calculation is being repeated several times for $25$ to $10,000$ simulated networks used in the likelihood approximation.

@@

implying that the smaller the RMSE, the more accurate is the estimator. Since the MCMLE has higher efficiency and converges to the MLE, the RMSE decreases as the number of simulated networks used for the likelihood approximation increases. On the other hand, the RMSE of the MPLE is a constant value since no network simulations are required. In order to compare the RMSE of the two estimation techniques, we take the log of the ratio of the MCMLE to the MPLE. As a result, a negative value indicates a better MCMLE performance, while a positive value indicates a better MPLE performance.\\[0.3cm] 
Figure \ref{rmse} visualizes the results of the simulation study. The solid line illustrates the results of the log relative RMSE of the Faux Mesa High network, while the dashed line illustrates the corresponding results of the Faux Magnolia High network. 

@@

\noindent The plots support the fact that larger networks require a larger sample size of simulated networks for the MCMLE to outperform the MPLE. While the fairly small Faux Mesa High network only requires a sample size of about $50-100$ networks, the larger Faux Magnolia High network requires a sample size of at least 1,500 networks for the MCMLE to surpass the MPLE.  For especially large networks (e.g., social media data) the sample size has to be set in order to justify the approximately exact, but computationally expensive, MCMLE method, may be prohibitive.

\section{Bootstrapped MPLE}

\noindent As discussed in the previous section, the MPLE converges to the MLE as the size of the network increases. Moreover, the MPLE is able to outperform the MCMLE if the sample size used in MCMLE is not large enough. The main reason why the MCMLE is still widely preferred is that, in contrast to the MPLE, it does not underestimate the standard errors (van Duijn et al. \cite{vanDuijnetal2009}). By the definition of the ERGM it is obvious that this model is an exponential family distribution where $\theta$ is the natural parameter and $\Gamma (G)$ is the sufficient statistic. For exponential family distributions, it is known that the sampling distribution of the MLE is multivariate normal with mean vector equal to the MLE and a covariance matrix equal to the inverse of the negative Hessian matrix $[-H]^{-1}$ of the likelihood function at the MLE. The problem with the MPLE is that calculating $[-H]^{-1}$ by the pseudolikelihood function will underestimate the variance of the MPLE \cite{vanDuijnetal2009}, resulting in an underestimate of the width of the confidence intervals. van Duijn et al. show that constructing 95\% MPLE confidence intervals can result in intervals that comprise the true value in less than 75\% instead of the nominal 95\%. In this paper, we are going to refer to the MPLE confidence intervals as \textit{logistic regression confidence intervals} simply because the MPLE is calculated using logistic regression methods that also use the inverse of the negative Hessian matrix as an estimate for the covariance matrix. \\[0.3cm]
\noindent Since the MPLE has the advantage of being approximately exact and computationally inexpensive, but has the disadvantage of underestimating corresponding confidence intervals, we apply a technique referred to as bootstrap resampling \cite{efron1982jackknife}. Bootstrap resampling refers to constructing a sampling distribution for the parameter estimate by resampling the data with replacement, and re-estimating the model on the resampled data. Under non-parametric bootstrap resampling, the data are resampled directly from the dataset. Under the parametric boostrap, the data is resampled from the estimated model. The idea of using boostrap resampling with MPLE for ERGM was first introduced by Desmarais and Cranmer \cite{Desmarais.2012} and provides a consistent estimate of MPLE confidence intervals. Desmarais and Cranmer argue that the MPLE is a multivariate \textit{M}-estimator (see Huber \cite{Huber1981}), a special class of robust estimators, meaning that bootstrap resampling consistently estimates the confidence intervals of the MPLE. \\[0.3cm]
\noindent However, the methods introduced by Desmarais and Cranmer \cite{Desmarais.2012} only applied to cases in which the researcher had a large sample of networks (e.g., a time series of networks), since the method they proposed was a non-parametric bootstrap. The non-parametric bootstrap cannot be applied when there is just a single network under study.

@@

This means that a relative computing time greater than 1 indicates that the MCMLE computing time is shorter, while a relative computing time smaller than 1 indicates that the bootstrapped MPLE provides faster results. \\[0.3cm]

>>>>

\begin{abstract}
With the growth of interest in network data across fields, the Exponential Random Graph Model (ERGM) has emerged as the leading approach to the statistical analysis of network data. ERGM parameter estimation requires the approximation of an intractable normalizing constant. Simulation methods represent the state-of-the-art approach to approximating the normalizing constant, leading to estimation by Monte Carlo maximum likelihood (MCMLE). MCMLE is accurate when a large sample of networks is used to approximate the normalizing constant. However, MCMLE is computationally expensive, and may be prohibitively so if the size of the network is on the order of 1,000 nodes (i.e., one million potential ties) or greater. When the network is large, one option is maximum pseudolikelihood estimation (MPLE). The standard MPLE is simple and fast, but generally underestimates standard errors. We show that a resampling method---the parametric bootstrap---results in accurate coverage probabilities for confidence intervals. We find that bootstrapped MPLE can be run in 1/5th the time of MCMLE. We study the relative performance of MCMLE and MPLE with simulation studies, and illustrate the two different approaches by applying them to a network of bills introduced in the United State Senate.

@@

\section{Introduction}
The field of network science faces a double-edge sword when it comes to computationally intensive research. First, the availability of digital source data has led the growth in network science to be synonymous with the growth in research on big data.  Faraj et al. (2008, pp. 19--20) \cite{faraj2008electronic} notes that, \begin{quote} ``In recent years the proliferation of advanced Information Technology has not only facilitated the collection of large-scale network data but also increased the availability of large-scale network data analysis tools and techniques. This has led to an explosion of interest in large network research in fields spanning biology, physics, mathematics, and social sciences.''  \end{quote} Second, analytical methods are growing more sophisticated, increasingly involving iterative and/or simulation-based optimization, rather than simple descriptive calculations \cite{snijders2011statistical}.  Closing the gap in terms of the size of the networks to which it is feasible to apply the most sophisticated methods of network modeling requires research into scalable methods of inference. We propose a method of statistical inference for one of the most popular models for networks---the exponential random graph model (ERGM), in which both parameter estimates and confidence intervals are derived, that can require less than half the compute time of currently used methods.

\section{The Exponential Random Graph Model}
The ERGM is a probabilistic model for networks \cite{chatterjee2013estimating,Wasserman.1996,robins.pattison.2007}. They can be used for link prediction \cite{lu2010supervised}, simulating network adjacency matrices \cite{hackney2006agent}, and testing theories regarding the processes underlying tie formation \cite{goodreau2009birds}. The ERGM was first introduced by Holland and Leinhardt (1981) \cite{holland1981exponential}. However, due to the intractable normalizing constant in the likelihood function of the ERGM, it did not see widespread and complete use until the 2000s, following the development of algorithms and software for efficient simulation-based methods for working with ERGM \cite{snijders2002markov}. Training ERGM using simulation-based methods is computationally expensive, and can still be prohibitively burdensome with data on big networks. Approximate methods of estimation, which are much more feasible with large networks, have existed for some time, but these methods perform poorly when it comes to characterizing the uncertainty in parameter estimates, which is necessary when assessing risk in predictions or simulation, or in hypothesis testing.\\
\indent The ERGM takes the adjacency matrix of an observed network $G^{obs}$, which is a matrix-valued random variable. This means that a network of $N$ nodes can be defined as a adjacency matrix $G=(g_{ij})\in \{0,1\}^{(N \times N)}$, where $g_{ij} \in \{0,1\}$ for all $i,j \in \{1,\dots , N\}$. $g_{ij}=1$ means that there is an edge between actors $i$ and $j$, while $g_{ij}=0$ indicates that these actors are not directly connected. Since the model does not consider loops, one has $g_{ii}=0$ for all $i \in \{1,\dots , N\}$. Furthermore, define
$$ \mathcal{G}(N) := \left\{ G \in \{0,1\}^{(N \times N)}: g_{ij} \in \{0,1\},~g_{ii}=0\right\}$$
as the set of all possible networks on $N$ nodes without loops. Note that the cardinality of set $\mathcal{G}(N)$ is increasing exponentially for every newly included actor, which results in $2^{N(N-1)/2}$ total elements. For even a small number of nodes, the cardinality of $\mathcal{G}(N)$ turns out to be an astronomically large number. For this reason calculating the likelihood function of the ERGM, which requires evaluating a normalizing constant on $\mathcal{G}(N)$  is either extremely time-consuming or with today's technology not achievable. As a consequence, many approximation methods have been provided by the literature, with the most popular method making use of Markov Chain Monte Carlo (MCMC) methods \cite{hunter2012computational}, as we will introduce in the next section.\\  
The probability function for the ERGM is defined as
\begin{equation}
\mathbb{P}_{\theta}(G)=\dfrac{\exp(\theta^T \cdot \Gamma(G))}{\sum_{G^* \in \mathcal{G}(N)} \exp(\theta^T \cdot \Gamma(G^*))} 
\label{ERGM}
\end{equation}
%
where $\theta \in \mathbb{R}^q$ is a $q-$dimensional vector of parameters, $\Gamma:\mathcal{G}(N) \to \mathbb{R}^q~,~G \mapsto (\Gamma_1(G),\dots,\Gamma_q(G))^T$ is a $q$-dimensional function of different network statistics and $c(\theta):= \sum_{G^* \in \mathcal{G}(N)} \exp(\theta^T \cdot \Gamma(G^*))$ is a normalization constant which ensures that (\ref{ERGM}) defines a probability function on $\mathcal{G}$.
%
As already mentioned, a specific network $G$ can be considered as a manifestation of a matrix-like random variable, whose probability of occurrence can be modeled with equation (\ref{ERGM}). The generative processes captured by a model (e.g., density, reciprocity, popularity, clustering) are informed by the decision regarding which network statistics (i.e., $\Gamma(\cdot)$) are incorporated (see Snijders et al. \cite{SnijdersTomA.B..2006} for a detailed discussion). The flexibility of the ERGM in capturing virtually any network generative process has led to it being applied broadly across several fields, including (but not limited to) sociology \cite{smith2016ethnic,srivastava2011culture}, economics \cite{lomi2012networks}, political science \cite{Cranmer.2011,lazer2010coevolution}, ecology \cite{dey2014individual,bouskila2015similarity}, and neuroscience \cite{simpson2011exponential,simpson2012exponential}.

\section{Estimation}
As mentioned above calculating $c(\theta)$ is not achievable for most cases with today's technology.  Straightforward application of either maximum likelihood estimation and Bayesian inference are, therefore, not possible. The first method proposed in the literature for estimating ERGM parameters was maximum pseudolikelihood estimation \cite{StraussIkeda1990}. Under maximum pseudolikelihood estimation (MPLE), the joint distribution is replaced by the product over conditional distributions \cite{besag1986statistical}.  
%
\begin{equation*}
\mathbb{P}_{\theta}(G) \approx \prod_{ij} \mathbb{P}_{\theta}\left(g_{ij}|G_{-ij}\right), 
%\label{ILogit}
\end{equation*}
where $G_{-ij}$ is the adjacency matrix, excluding element $ij$. The conditional probability of a tie in ERGM reduces, conveniently, to a logistic regression form given by 
\begin{equation*}
\mathbb{P}_{\theta}\left(g_{ij}=1| G_{-ij} \right) = \text{\\} \text{logit}^{-1}\left( \theta^T \cdot \delta(\Gamma(G))  \right), 
%\label{ILogit}
\end{equation*}
where $\delta(\Gamma(G))$ is the ``change statistic'' given by the difference in the network statistics when the $ij$ element is toggled from 0 to 1 (i.e., $\Gamma(G|g_{ij}=1))-\Gamma(G|g_{ij}=0)$), and $\text{logit}^{-1}(x)=1/(1+\exp(-x))$ \cite{goodreau2009birds}. For the ERGM, the pseudolikelihood function can be maximized using logistic regression software, in which the dependent variable is given by the elements of the adjacency matrix, and the covariates are given by the values of the change statistics corresponding to each element of the adjacency matrix.

@@

The contemporary conventional approach to estimating $\theta$, introduced by Snijders \cite{snijders2002markov}, is based on a Markov Chain Monte Carlo (MCMC) approximation of the MLE.\\
This Monte Carlo maximum likelihood method (MCMLE) is based on a more direct attempt to approximate the  log-likelihood function derived from (\ref{ERGM}). The log-likelihood function is not evaluated directly, rather, the log ratio of the likelihood under a proposed value of the parameters $\theta$, and an initial value of the parameters $\theta_0$, is approximated using $L$ networks simulated from the ERGM with parameter values $\theta_0$. The approximation, detailed in Snijders (2002) \cite{snijders2002markov} is given by 
%
\begin{eqnarray*}
&&\text{loglik}(\theta)-\text{loglik}(\theta_0)\\&=&- \log(c(\theta))+\log(c(\theta_0))\\
                                             &=&- \log \left( \frac{c(\theta)}{c(\theta_0)} \right)\\
                                             &=&- \log \left( \mathbb{E}_{\theta_0}\left[ \exp\left((\theta - \theta_0)^T \cdot \Gamma(Y)\right) \right] \right)\\
                                             &\approx &- \log \left( \frac{1}{L} \cdot \sum_{i=1}^{L}  \exp \left((\theta - \theta_0)^T \cdot \Gamma(G_i) \right) \right).
\end{eqnarray*}
By differentiating this equation on both sides with respect to $\theta$ one gets an approximate score function:
%
\begin{equation}
s(\theta) \approx -\frac{\partial}{\partial \theta} \log \left( \frac{1}{L} \cdot \sum_{i=1}^{L}  \exp \left((\theta - \theta_0)^T \cdot \Gamma(G_i) \right) \right)
\label{score}
\end{equation}
%
This approximate score function now can be used as usual, i.e., it can be iteratively approximately optimized with the \textit{Newton-Raphson algorithm} or by \textit{Fisher scoring}.

MCMC methods are used to simulate the $L$ networks. As we demonstrate below, the MCMLE grows more accurate as $L$ increases. Indeed, MCMLE approaches the MLE as the number of networks simulated goes to infinity. Snijders (2002) \cite{snijders2002markov} provides a Metropolis Hastings algorithm to simulate networks:
Choose a matrix $G^{(0)} \in \mathcal{G}(N)$ to start with (e.g., the observed network). For $k \in \{0,...,L-1\}$ recursively proceed as follows:\\
\begin{enumerate}
\item Randomly choose an edge $(i,j)$ where $i \neq j$ from $G^{(k)}$
\item Compute the value
$$\pi := \dfrac{\mathbb{P}_{\theta}(Y_{ij} \neq g_{ij}^{(k)} | Y_{ij}^c=G_{ij}^c)}{\mathbb{P}_{\theta}(Y_{ij}= g_{ij}^{(k)} | Y_{ij}^c=G_{ij}^c)}$$
\item Fix $\delta:= \min\{1, \pi\}$ and draw a random number $Z$ from Bin$(1, \delta)$. If
\begin{itemize}
\item $Z=0$, let $G^{(k+1)} := G^{(k)}$ 
\item $Z=1$, define $G^{(k+1)}$ via
$$g_{pq}^{(k+1)}=\begin{cases}
1-g_{pq}^{(k)}& \text{if}~ (p,q)=(i,j) \\
g_{pq}^{(k)} &\text{if}~ (p,q) \neq (i,j) 
\end{cases}$$
\end{itemize}
\item Start at step 1 with $G^{(k+1)}$.
\end{enumerate}
\vspace{0.3cm}
The depicted algorithm provides a sequence of random networks $G_(0),\dots,G_(L)$ via a Metropolis-Hastings sampler \cite{chib1995understanding}. Since the original matrix was chosen randomly and the first simulated networks are very dependent on the chosen matrix (only one edge is changed per iteration), usually
the first $B$ networks, where $N << B << L$, are discarded as the so called Burn-In.

\section{Efficiency of MPLE and MCMLE}
As mentioned in the previous section the MPLE approaches the MLE as the size of the networks increase and as a consequence, is a consistent estimator (see Lindsay \cite{Lindsay1988}, Strauss and Ikeda \cite{StraussIkeda1990}, Hyvarinen \cite{Hyvarinen2006}, Desmarais and Cranmer \cite{Desmarais.2012}). This implies that for an increasing number of nodes, the MPLE converges in probability to the MLE, meaning that for large enough networks the MPLE performs as well as or better than MCMLE, and requires less compute time. At this point we want to mention that we are familiar with the work of Shalizi and Rinaldo \cite{shalizi2013}, arguing that consistency is not given in the ERGM framework. They prove that one cannot run an ERGM on a sub-network in order to make inferences about the full network.  The way we use the term \textit{consistency} in this paper is different and aligns with the way consistency is defined by Lindsay \cite{Lindsay1988}, i.e. instead of considering sub-networks that converge to the full size network, we argue that both, the MLE as well as the MPLE, approach the true coefficient values as the size of networks generally increases.

To illustrate the relative efficiency of MPLE and MCMLE we run a simulation study. Desmarais and Cranmer \cite{Desmarais.2012} show the MPLE outperforms the MCMLE if the number of simulated networks used to approximate the likelihood in MCMLE is not large enough. It is even more remarkable that the number of simulated networks needed for the MCMLE, in order to surpass the MPLE increases as the size of the network (i.e., the number of nodes in the network) increases. This means that, for very large networks, it becomes difficult to determine the number of simulated networks required for the MCMLE to outperform the MPLE. In other words, the larger the network, the more computationally intensive it becomes to use MCMLE in a way that out-performs MPLE.\\
\indent To demonstrate this disadvantage of the MCMLE we conduct a simulation study using Goodreau's \cite{HunterDavidR..2008} Faux Mesa High School data, which represents a simulation of an in-school friendship network among 203 students as well as the Faux Magnolia High School data, representing an in-school friendship network among 1451 students. The data for both networks originates from Resnick et al. \cite{Resnicketal1997}.\\
\indent For both networks, we first calculate the MCMLE and treat the estimated coefficients as the network's true values $\theta$. Then, we take the same parametrization, using the number of edges, the nodal attribute for gender, and the geometrically weighted edgewise shared partners (gwesp) distribution (see Hunter \cite{Hunter.2006}) where we fix the decay parameter $\lambda$ at $0.25$.

@@

\indent We simulate $m=500$ new networks using the 'true' coefficients and estimate the MPLE as well as the MCMLE of these simulated networks. For every single simulated network the MCMLE calculation is being repeated several times for $25$ to $10,000$ simulated networks used in the likelihood approximation.

@@

implying that the smaller the RMSE, the more accurate is the estimator. Since the MCMLE has higher efficiency and converges to the MLE, the RMSE decreases as the number of simulated networks used for the likelihood approximation increases. On the other hand, the RMSE of the MPLE is a constant value since no network simulations are required. In order to compare the RMSE of the two estimation techniques, we take the log of the ratio of the MCMLE to the MPLE. As a result, a negative value indicates a better MCMLE performance, while a positive value indicates a better MPLE performance.\\
\indent Figure \ref{rmse} visualizes the results of the simulation study. The solid line illustrates the results of the log relative RMSE of the Faux Mesa High network, while the dashed line illustrates the corresponding results of the Faux Magnolia High network. 

@@

The plots support the fact that larger networks require a larger sample size of simulated networks for the MCMLE to outperform the MPLE. While the fairly small Faux Mesa High network only requires a sample size of about $50-100$ networks, the larger Faux Magnolia High network requires a sample size of at least 1,500 networks for the MCMLE to surpass the MPLE.  For especially large networks (e.g., social media data) the sample size has to be set in order to justify the approximately exact, but computationally expensive and potentially prohibitive MCMLE method.

\section{Bootstrapped MPLE}
As discussed in the previous section, the MPLE converges to the MLE as the size of the network increases. Moreover, the MPLE is able to outperform the MCMLE if the sample size used in MCMLE is not large enough. The main reason why the MCMLE is still widely preferred is that, in contrast to the MPLE, it does not underestimate the standard errors (van Duijn et al. \cite{vanDuijnetal2009}). By the definition of the ERGM it is obvious that this model is an exponential family distribution where $\theta$ is the natural parameter and $\Gamma (G)$ is the sufficient statistic. For exponential family distributions, it is known that the sampling distribution of the MLE is multivariate normal with mean vector equal to the MLE and a covariance matrix equal to the inverse of the negative Hessian matrix $[-H]^{-1}$ of the likelihood function at the MLE. The problem with the MPLE is that calculating $[-H]^{-1}$ by the pseudolikelihood function will underestimate the variance of the MPLE \cite{vanDuijnetal2009}, resulting in an underestimate of the width of the confidence intervals. van Duijn et al. show that constructing 95\% MPLE confidence intervals can result in intervals that comprise the true value in less than 75\% instead of the nominal 95\%. In this paper, we are going to refer to the MPLE confidence intervals as \textit{logistic regression confidence intervals} simply because the MPLE is calculated using logistic regression methods that also use the inverse of the negative Hessian matrix as an estimate for the covariance matrix. \\
\indent Since the MPLE has the advantage of being approximately exact and computationally inexpensive, but has the disadvantage of underestimating corresponding confidence intervals, we apply a technique referred to as bootstrap resampling \cite{efron1982jackknife}. Bootstrap resampling refers to constructing a sampling distribution for the parameter estimate by resampling the data with replacement, and re-estimating the model on the resampled data. Under non-parametric bootstrap resampling, the data are resampled directly from the dataset. Under the parametric boostrap, the data is resampled from the estimated model. The idea of using boostrap resampling with MPLE for ERGM was first introduced by Desmarais and Cranmer \cite{Desmarais.2012} and provides a consistent estimate of MPLE confidence intervals. Desmarais and Cranmer argue that the MPLE is a multivariate \textit{M}-estimator (see Huber \cite{Huber1981}), a special class of robust estimators, meaning that bootstrap resampling consistently estimates the confidence intervals of the MPLE. \\
\indent However, the methods introduced by Desmarais and Cranmer \cite{Desmarais.2012} only applied to cases in which the researcher had a large sample of networks (e.g., a time series of networks), since the method they proposed was a non-parametric bootstrap. The non-parametric bootstrap cannot be applied when there is just a single network under study.

@@

This means that a relative computing time greater than 1 indicates that the MCMLE computing time is shorter, while a relative computing time smaller than 1 indicates that the bootstrapped MPLE provides faster results. \\
